Damegender: Writing and Comparing Gender Detection Tools

### 2020-05-14 brief.tex
brief.pdf, brief.tex: creation

\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{twocolceurws}

@@

\author{
David Arroyo Menéndez \\ GSYC \\ Universidad Rey Juan Carlos I \\ d.arroyome@alumnos.urjc.es
\and
Jesús González Barahona \\ GSYC \\ Universidad Rey Juan Carlos I \\ jgb@gsyc.es
}

\institution{}

\begin{document}
\maketitle

\begin{abstract}
The variable sex (male or female) is one of most used variables for
any study in sociology, but this variable can be hidden in Internet
communities. The gender detection from a name is an important problem
in Natural Language Processing to decide if a string labeled as name
is classified as male or female. An engineer will find useful
make gender detection from a name retrieving information from social
networks, mailing lists, instant messaging, software repositories,
papers, etc. To achieve gender equality and empower all women and
girls is a goal in sustanaible development in United Nations, so to
measure the gender gap is a previous step to find solutions to reduce
it.

@@

\section{Introduction}

@@

The problems addressed are:
\begin{itemize}
\item Evaluate quality/price with different commercial solutions.
\item Think about solutions using free licenses.
\item Treatment with names without census, for example, nicknames,
  diminutives, ...
\item Massive gender detection from Internet, for example, mailing
  lists, software repositories, ...
\end{itemize}

In this paper, these problems are faced writing a Python solution for:
\begin{itemize}
\item To evaluate quality of different solutions applying metrics
  suggested by ~\cite{10.7717/peerj-cs.156}
\item To understand the current technology in detail, I have developed
  a tool guessing gender from a name giving support to Spanish and
  English from the open data census provides by the states.
\item To fix the problem with nicknames and diminutives, we have
  developed a machine learning solution to strings not found in the
  census dataset.
\item To do proof-of-concept tests applying Perceval to detect
  gender in mailing lists and software repositories.
\end{itemize}

In Section 2, we explain the current solutions to the problems. In
Section 3, we present the results evaluating the current Application
Programming Interfaces with our software. In Section 4, we discuss
attempts and problems releasing with a free license a gender detection
from name program. In Section 5, we discuss how to obtain Open Datasets
counting names and gender. In Section 6, we describe our
machine learning solution. In Section 7, we describe general
implementation details. Finally, in Section 8 we summarize our
findings, and describe extensions to the work that we are currently
exploring.

\section*{3. Application Programming Interfaces}

\subsection*{Market}

We have reproduced to ~\cite{10.7717/peerj-cs.156} and updated on
27/06/2019 and we are showing the results in
~\ref{table:featuresAndGenderDetectionToolsByName}.

\begin{longtable}[]{@{}lllllll@{}}
\toprule
Feature & Gender API & genderguesser & genderize.io & NameAPI & NamSor & Damegender\tabularnewline
\midrule
\endhead
Database size & 431*10^6 & 48.528 & 114*10^6 & 1.428.345 & 4407*10^6  & 197.271\footnote{This cipher is the sum of names in United States of America, United Kingdom, Uruguay and Spain} \tabularnewline
Regular data updates & yes & no & yes & yes & yes & yes, dev\tabularnewline
Unstructured strings & yes & no & no & yes & no & yes\tabularnewline
Handles surnames & yes & no & no & yes & yes & yes\tabularnewline
Non-Latin alphabets & partially & no & partially & yes & yes & no\tabularnewline
Geo-localization & yes & no & no & yes & yes & no\tabularnewline
Exists locale & yes & yes & yes & yes & yes & yes\tabularnewline
Assingment type & probabilistic & binary & probabilistic & probabilistic & probabilistic & prob\tabularnewline
Free params & total, prob & gender & total, prob & confidence & scale & total, prob\tabularnewline
Guessing with ML & no & no & no & no & no & yes\tabularnewline
Free license & no & yes & no & no & no & yes\tabularnewline
API & yes & no & yes & yes & yes & future\tabularnewline
free requests limited & yes (200) & unlimited & yes (1000) & yes & yes & unlimited\tabularnewline
\bottomrule
\caption{Features and gender detection tools by name}
\label{table:featuresAndGenderDetectionToolsByName}
\end{longtable}

All solutions have increased the database size from
~\cite{10.7717/peerj-cs.156}. NameAPI and GenderAPI is reaching more
features. The tools with a free license have not many features, so for
now that will not be the trend in many situations. Today, one good
solution quality and price is Namsor, which provides unlimited names
through an Application Programming Interface with many features in the
task to detect gender from the name.

\subsection*{Reproducing accuracies and confusion matrix}

@@

\section*{Datasets}

\begin{verbatim}
$ python3 main.py David --total="ine"
David gender is male
363559  males for David from INE.es
0 females for David from INE.es
\end{verbatim}

In Damegender, we are including Open Data census about names and
gender, such as INE.es or USA and United Kingdom (births and dies). We
want datasets provided by the software package to increment the speed
retrieving data.

\section{Machine Learning}

@@

Take a look to the most informative features results with the
different datasets:
\begin{longtable}[]{@{}llllllll@{}}
  \toprule
Dataset & Char A & Last Char A & Last Char O & Last Char Consonant & Last Char Vocal & First Char Consonant & First Char Vocal & \tabularnewline
  \midrule
  \endhead
 Uruguay (F) \footnote{F is for females and M is for males} &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177 & \tabularnewline
 Uruguay (M) \footnotemark[3] &    0.643 &         0.249 &         0.062 &                 0.766 &             0.234 &                  0.771 &              0.228 & \tabularnewline
 Spain (F) \footnotemark[3]  &    0.922 &         0.588 &          0.03 &                 0.271 &             0.728 &                  0.772 &              0.228 & \tabularnewline
 Spain (M) \footnotemark[3]  &    0.818 &          0.03 &         0.268 &                 0.569 &              0.43 &                  0.763 &              0.236 & \tabularnewline
 UK (F) \footnotemark[3]     &    0.825 &         0.374 &         0.013 &                 0.322 &             0.674 &                  0.765 &              0.235 & \tabularnewline
 UK (M) \footnotemark[3]     &    0.716 &         0.036 &         0.039 &                  0.78 &             0.218 &                  0.799 &                0.2 & \tabularnewline
 USA (F) \footnotemark[3]    &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177 & \tabularnewline
 USA (M) \footnotemark[3]    &    0.643 &          0.02 &         0.061 &                 0.765 &             0.234 &                   0.84 &              0.159 & \tabularnewline
\bottomrule
\caption{Informative Features in Different Countries}
\label{table:InfoFeatures}
\end{longtable}

@@

 Support Vector Machines             &    0.879 &     0.972 &   0.972 &    1.0 & \tabularnewline
 Random Forest                       &    0.862 &     0.902 &   0.902 &    1.0 & \tabularnewline
 NLTK (Bayes)                        &    0.862 &     0.902 &   0.902 &    1.0 & \tabularnewline
 Multinomial Navie Bayes             &    0.782 &     0.791 &   0.791 &    1.0 & \tabularnewline
 Tree                                &    0.764 &     0.821 &   0.796 &    1.0 & \tabularnewline
 Stochastic Gradient Distribution    &    0.709 &     0.943 &   0.815 &    1.0 & \tabularnewline
 Gaussian Naive Bayes                &    0.709 &     0.968 &   0.887 &    1.0 & \tabularnewline
 Bernoulli Naive Bayes               &    0.699 &     0.965 &   0.816 &    1.0 & \tabularnewline
 AdaBoost                            &    0.698 &     0.965 &   0.815 &    1.0 & \tabularnewline
 Multi Layer Perceptron              &    0.677 &     0.819 &   0.755 &    1.0 & \tabularnewline

@@

\section{State of Art}

\subsection{Comparing Commercial Solutions}

~\cite{10.7717/peerj-cs.156} are proposing a good metrics set to
classify these commercial Application Programming Interfaces (features,
measuring errors and success, ...). The features observed are:
Database size (January 2018), Regular data updates, Handles
unstructured full name strings, Handles surnames, Handles non-Latin
alphabets, Implicit geo-localization, Assignment type, Free
parameters, Open source, Application Programming Interface, Monthly
free requests, Monthly subscription cost (100,000 requests/month)).

In the commercial tools is being used different ways to express
probability (confidence, scale, accuracy, precision, recall, ...).

\subsubsection{Datasets}

In ~\cite{berners2001semantic} a world was envisioned where public
structured data could be interconnected with software agents to
process these data, perhaps only recovering information, but mixed
with distributed artificial intelligence would give a big jump to the
semantic richness to the web.

~\cite{janssen2012benefits} shows serious profits for the states
adopting Open Data in three categories (1) political and social, (2)
economical, (3) operational and technical. So, Open Data is a
breakthrough towards the Semantic Web.

We can find Open Data about names and gender in census of citizens in
states and commercial solutions. Free software packages such as
~\cite{krawetz2006gender} or ~\cite{loper2002nltk} is providing good
datasets about names and gender. So, Damegender incorporates different
lists of names from free software solutions wrote before (Natural
Language ToolKit, Gender Guesser, ...) and from Open Data census
(United Kingdom, USA, Spain, Uruguay, ...).

Wikidata ~\cite{42240} provides a semantic and open database about
Wikipedia allowing retrieve information with Sparql, such as names and
gender.

~\cite{10.7717/peerj-cs.156} describes different ways to build a
dataset on hand looking for names in papers, scientific websites,
wikipedia, biographies, photos, ...)

\section*{Implementation}

@@

\section*{Conclusions}

@@

@@

%\bibliographystyle{alpha} 
%% \bibliography{satbibtex}

%% \bibitem[Com79]{Comer-btree}
%% D.~Comer.
%% \newblock The ubiquitous b-tree.
%% \newblock {\em Computing Surveys}, 11(2):121--137, June 1979.

%% \bibitem[Knu73]{Knuth-vol3}
%% D.~E. Knuth.
%% \newblock {\em The Art of Computer Programming -- Volume 3 / Sorting and
%%   Searching}.
%% \newblock Addison-Wesley, 1973.

%% \bibliographystyle{alpha} 
%% \bibliography{bibtex}
%inline the .bbl file directly for mailing to authors.

>>>>

%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

%\documentclass[fleqn,10pt]{llncs} % for journal submissions
\documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions

@@

\author[1]{David Arroyo Menéndez}
\affil[1]{Universidad Rey Juan Carlos}

\keywords{Gender gap, Gender detection tools, Software repositories}
\begin{abstract}
The variable sex (male or female) is one of most used variables for
any study in sociology, but this variable can be hidden in Internet
communities. The gender detection from a name is an important problem
in Natural Language Processing to decide if a string labeled as name
is classified as male or female. An engineer will find useful
make gender detection from a name retrieving information from social
networks, mailing lists, instant messaging, software repositories,
papers, etc. To achieve gender equality and empower all women and
girls is a goal in sustanaible development in United Nations, so to
measure the gender gap is a previous step to find solutions to reduce
it.

@@

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{1. Introduction}

@@

The problems addressed are:
\begin{itemize}[noitemsep]
\item Evaluate quality/price with different commercial solutions.
\item Think about solutions using free licenses.
\item Treatment with names without census, for example, nicknames,
  diminutives, ...
\item Massive gender detection from Internet, for example, mailing
  lists, software repositories, ...
\end{itemize}

In this paper, these problems are faced writing a Python solution for:
\begin{itemize}[noitemsep]
\item To evaluate quality of different solutions applying metrics
  suggested by ~\cite{10.7717/peerj-cs.156}
\item To understand the current technology in detail, I have developed
  a tool guessing gender from a name giving support to Spanish and
  English from the open data census provides by the states.
\item To fix the problem with nicknames and diminutives, we have
  developed a machine learning solution to strings not found in the
  census dataset.
\item To do proof-of-concept tests applying Perceval to detect
  gender in mailing lists and software repositories.
\end{itemize}

\section*{2. State of Art}

\subsection*{Comparing Commercial Solutions}

~\cite{10.7717/peerj-cs.156} are proposing a good metrics set to
classify these commercial Application Programming Interfaces
(features, measuring errors and successes, ...). The features observed
are: Database size (January 2018), Regular data updates, Handles
unstructured full name strings, Handles surnames, Handles non-Latin
alphabets, Implicit geo-localization, Assignment type, Free
parameters, Open source, Application Programming Interface, Monthly
free requests, Monthly subscription cost (100,000 requests/month)).

It's possible different ways to express probability about the
succesful (confidence, scale, accuracy, precision, recall, ...),
confusion matrix to understand where it's successes or fails and
differente errors measures (error coded, error coded without not
applicable values, error gender bias, not applicable coded).

\subsection*{Datasets}

In Damegender, we are including Open Data census about names and
gender, such as INE.es, Uruguay, USA and United Kingdom. We want
datasets provided by the software package to increment the speed
retrieving data.

Some Open Datasets, such Spain (INE.es), or United States of America
is providing support to surnames and the relationship with the
ethnicity. For example, in United States of America is giving a
probability with the race and in Spain INE.es is giving the number of
people with a surname with a nationality different to the spanish
nationality. The propietary solutions is allowing infer the gender
with the surname as another parameter, due to names such as Andrea
changes the gender depending the nationality: in Italia would be male,
but in Spain would be female.

\section*{Reproducing accuracies and confusion matrix}

@@

\section{5. Machine Learning}

@@

Take a look to the most informative features results with the
different datasets:
\begin{longtable}[]{@{}llllllll@{}}
  \toprule
Dataset & char a & last char a & last char o & last char consonant & last char vocal & first char consonant & first char vocal & \tabularnewline
  \midrule
  \endhead
 Uruguay (F) \footnote{F is for females and M is for males} &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177 & \tabularnewline
 Uruguay (M) \footnotemark[3] &    0.643 &         0.249 &         0.062 &                 0.766 &             0.234 &                  0.771 &              0.228 & \tabularnewline
 Spain (F) \footnotemark[3]  &    0.922 &         0.588 &          0.03 &                 0.271 &             0.728 &                  0.772 &              0.228 & \tabularnewline
 Spain (M) \footnotemark[3]  &    0.818 &          0.03 &         0.268 &                 0.569 &              0.43 &                  0.763 &              0.236 & \tabularnewline
 UK (F) \footnotemark[3]     &    0.825 &         0.374 &         0.013 &                 0.322 &             0.674 &                  0.765 &              0.235 & \tabularnewline
 UK (M) \footnotemark[3]     &    0.716 &         0.036 &         0.039 &                  0.78 &             0.218 &                  0.799 &                0.2 & \tabularnewline
 USA (F) \footnotemark[3]    &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177 & \tabularnewline
 USA (M) \footnotemark[3]    &    0.643 &          0.02 &         0.061 &                 0.765 &             0.234 &                   0.84 &              0.159 & \tabularnewline
\bottomrule
\caption{Informative Features in Different Countries}
\label{table:InfoFeatures}
\end{longtable}

@@

 Support Vector Machines             &    0.879 &     0.972 &   0.972 &    1.0 & \tabularnewline
 Random Forest                       &    0.862 &     0.902 &   0.902 &    1.0 & \tabularnewline
 NLTK (Bayes)                        &    0.862 &     0.902 &   0.902 &    1.0 & \tabularnewline
 Multinomial Navie Bayes             &    0.782 &     0.791 &   0.791 &    1.0 & \tabularnewline
 Tree                                &    0.764 &     0.821 &   0.796 &    1.0 & \tabularnewline
 Stochastic Gradient Distribution    &    0.709 &     0.943 &   0.815 &    1.0 & \tabularnewline
 Gaussian Naive Bayes                &    0.709 &     0.968 &   0.887 &    1.0 & \tabularnewline
 Bernoulli Naive Bayes               &    0.699 &     0.965 &   0.816 &    1.0 & \tabularnewline

@@

\section*{6. Implementation}

@@

\section*{7. Conclusions}

@@

\bibliography{bibtex}

@@
